Fit by design: Developing substrate-specific seed mixtures for functional dike restoration

### 2023-02-23 manuscript.Rmd
Update manuscript.Rmd

Dikes are promising sites for the restoration of species-rich grasslands because they can increase habitat area and connectivity of semi-natural grasslands and therefore significantly contribute to biodiversity conservation in agricultural landscapes [@Batori.2020]. Steep slopes with different exposition, contrasting substrate layers and dense swards for erosion protection characterize dikes [@Batori.2016; @berendse2015]. Dikes can reconcile several requirements [@teixeira2022], which can be fostered by an adapted seed--substrate combination.

@@

Under the substrate, a 5 cm thick drainage layer of gravel (0-16 mm) was installed. Soil samples of the three substrates were tested on both expositions by mixing several sub-samples from different plots. The sand admixture changed the soil texture, increased the C/N ratio, reduced Calciumcarbonat, but did hardly changed the quite high pH values (Table \@ref(tab:substrate)). Phosphate and potassium were rather scarce for agricultural soils and magnesium high [@lfl2022]. The vegetation was surveyed in June or July 2018--2021 [@BraunBlanquet.1964] and Londo scale was used [@londo1976]. The establishment rates of species are recorded in Appendix A4. In total, 48 species of the species pool of hay meadows (87%) and 46 (79%) of dry grasslands) established by 2021. The species established 31 ± 22% of their sown plot. In total, 274 species were detected (Appendix A5).

@@

To measure the effects of the treatments on our three response variables, we calculated Bayesian linear mixed-effects models (BLMM) with the random effect plot nested in block with the Cauchy prior [see @lemoine2019]. Furthermore, we included as a fixed effect the botanists recording a certain plot. For the simple effects of the treatments, we chose plausible weakly informative priors. To evaluate the influence of the priors, prior predictive checks and models with non-informative priors were calculated. For the computation, we used four chains, a thinning rate of two, 5,000 iterations for warm-up, and 10,000 in total. For evaluating the computation, MCMC chain convergence was checked with trace plots and by the R-hat values, and MCMC chain resolution by the effective sampling size (ESS). Posterior predictive checks were done with Kernel density estimates histograms of statistics skew and leave-one-out (LOO) cross-validation [see @gabry2019]. Finally, models were compared with the Bayes factor (BF) and Bayesian *R*² values [@gelman2019]. A non-metric multidimensional scaling ordination (NMDS) with Jaccard dissimilarity was used to visualize variation in species composition in space and time (2D-stress 0.21). Seven species were excluded because they had an accumulated cover over all plots of less than 0.5%. In total, 343 species were included in the ordination.

@@

Restoration on agricultural soils can have limited success due to high nutrient loads [@walker2004] but can lead to positive results, if mixed with mineral material as sand and gravel [@chenot-lescure2022]. As expected, sand admixture reduced nutrient loads and led to higher persistence of sown species while 50% admixture did not further increase the effect. In addition, the effect only appeared in north exposition and the effect size of about 6% in the fourth year after restoration was rather small. Substrate depth did not significantly affect persistence or Favorable Conservation Status (FCS) similarly to earlier studies [@baer2004]. Larger differences in soil depths might be necessary to observe negative effects by thicker substrate layers as was shown for prairies [@dornbush2010]. Neither, seed density had an effect on persistence and FCS which fits to results of @Kaulfuss.2022, who found out that a certain amount of seeds are necessary for a successful establishment of target species but higher densities do not further improve the result and have rather a slight negative effect.

The vegetation in south exposition had a different species composition which confirms the findings of @Batori.2016 in Hungary. Here, the differences might have some unwanted experimental reasons: the geotextile, which had been implemented on the southern slope, was removed after two weeks. This led to the drying of many seedlings which was amplified by the intense drought of 2018 and 2019 [@hari2020]. This drought could have decreased germination and establishment of sown species also in general because a moist spring with a mesic temperature improves establishment [@Larson.2021; @orrock2023]. The interesting point here is the lasting negative effect on persistence and FCS which suggests a legacy effect by adverse weather conditions after sowing as it was observed by other studies [@Stuble.2017; @Groves.2020]. The problems during the establishment phase might have led to an alternative trajectory `r word_spec("(reference?)")`<!--# Add ref--> and could probably only be met by re-sowing. The drying of seedlings might also have leveled the distinction of the seed mixture types on the south exposition.

@@

Our aim was to identify the best seed--substrate combination in respect of restoration effectiveness and biodiversity. For evaluating effectiveness, we measured the persistence of the sown species and for fostering plant biodiversity in general, we measured the FCS. Though, we could not identify an interaction effect for neither of the indices. We would have expected a better performance of hay-meadow seed mixtures with lower sand admixture and for dry grasslands with higher sand admixture. Our results suggest that, at least after four years, the substrate conditions are within the range of both seed mixture types (hay meadows, dry grasslands). Although, both types are clearly plant-sociologically and functionally distinct, they are still relative closely related why they contain shared species and could develop on similar site conditions [Appendix A3, @Oberdorfer.1993; @Oberdorfer.1993b]. Other grassland studies could identify more or less clear interactions of opposing habitat preferences or functional traits along the gradients of productivity, moisture, and nutrients [@Zirbel.2020; @Freitag.2021; @Kaulfuss.2022]. However, these studies did not work with an experimental set up of different seed--substrate combinations, but analyzed the result of habitat and biotic filtering after one, five and about 15 years, respectively. Furthermore, the non-existence of an ideal combination could be explained by priority effects which means that the species of the non-perfect seed mixture could establish earlier and preempted the available niches for the species of related habitat types [@Fukami.2015].

@@

Local setting and design of the multifactorial experiment on grassland sowing at dikes. The experiment was located on a dike at river Danube in SE Germany. The 288 plots were allocated in six blocks (white squares on upper photo) and on the north and south slope (central photo) [Both aerial photos: @bayerischevermessungsverwaltung2023]. Four treatments were conducted: sand admixture, substrate depth, seed density, and seed mixture types H and D. The western half of a block was with a shallow substrate depth and within this half the substrates had different sand admixtures. The photo on the bottom shows the northern slope of one block in 2021, four years after sowing (Photo: Markus Bauer).

@@

```{r fig.id = "persistence", fig.asp = 0.4848, fig.width = 6.49, fig.cap = "Persistence of sown species of sown grassland communities at a river dike. The plots had substrates with different sand admixtures and were sown with different seed mixture types. Per plot 20 species were sown. Shown are the medians and 95% credible intervals of the posterior distributions which were derived from a Bayesian linear mixed-effects model (BLMM)."}
knitr::include_graphics(
  here::here("outputs", "figures", "figure_3_persistence_epred_800dpi_16.5x8cm.tiff")
  )
```

@@

```{r fig.id = "fcs", fig.asp = 0.4848, fig.width = 6.49, fig.cap = "The Favorable Conservation Status (FCS) of sown grassland communities at a river dike. The plots had substrates with different sand admixtures and were sown with different seed mixtures. The zero line indicate that target and non-target species are balanced. Positive values indicate that there are more target species. Shown are the medians and 95% credible intervals of the posterior distributions which were derived from a Bayesian linear mixed-effects model (BLMM)."}
knitr::include_graphics(
  here::here("outputs", "figures", "figure_4_fcs_epred_800dpi_16.5x8cm.tiff")
  )
```

@@

```{r fig.id = "recovery", fig.asp = 0.4848, fig.width = 6.49, fig.cap = "The recovery or successional distance (*d~jt,0~*) of sown grassland communities at a river dike in comparison to the respective reference sites (hay meadows or dry calcareous grasslands). The plots had substrates with different sand admixtures and were sown with different seed mixtures. The zero lines indicate the mean position of the reference sites for each habitat type on the NMDS axis 1 (Figure 7). The grey area marks the standard deviation of the position of the reference sites. Shown are the medians and 95% credible intervals of the posterior distributions which were derived from a Bayesian linear mixed-effects model (BLMM)."}
knitr::include_graphics(
  here::here("outputs", "figures", "figure_5_recovery_epred_800dpi_16.5x8cm.tiff")
  )
```

@@

NMDS ordination based on Sørensen dissimilarity of the species composition of 288 plots on dike grasslands over four years after sowing the seed mixtures (black squares). In addition, 98 positive and negative reference plots were included (filled symbols) from older dike grasslands in the surroundings [@bauer2023] and six plots from sPlotOpen [@Sabatini.2021]. The circles show the standard error of the groups. 2D stress: 0.21.

>>>>

Dikes are promising sites for the restoration of species-rich grasslands because they can increase habitat area and connectivity of semi-natural grasslands and therefore significantly contribute to biodiversity conservation in agricultural landscapes [@Batori.2020]. Steep slopes with different exposition, contrasting substrate layers and dense swards for erosion protection characterize dikes [@Batori.2016; @berendse2015; @husicka2003]. Dikes can reconcile several requirements [@teixeira2022], which can be fostered by an adapted seed--substrate combination.

@@

Under the substrate, a 5 cm thick drainage layer of gravel (0-16 mm) was installed. Soil samples of the three substrates were tested on both expositions by mixing several sub-samples from different plots. The sand admixture changed the soil texture, increased the C/N ratio, reduced Calciumcarbonat, but did hardly change the pH values which were in the weak alkaline range (Table \@ref(tab:substrate)). The pH values correspond to a well established dike vegetation on river Rhine [@husicka2003], phosphate and potassium were rather scarce for agricultural soils and magnesium high [@lfl2022]. The vegetation was surveyed in June or July 2018--2021 [@BraunBlanquet.1964] and Londo scale was used [@londo1976]. The establishment rates of species are recorded in Appendix A4. In total, 48 species of the species pool of hay meadows (87%) and 46 (79%) of dry grasslands established by 2021, which are rather good ratios [@hedberg2010]. The species established 31 ± 22% of their sown plot. In total, 274 species were detected (Appendix A5).

@@

To measure the effects of the treatments on our three response variables, we calculated Bayesian linear mixed-effects models (BLMM) with the random effect plot nested in block with the Cauchy prior [see @lemoine2019]. Furthermore, we included as a fixed effect the botanists recording a certain plot. For the simple effects of the treatments, we chose plausible weakly informative priors. To evaluate the influence of the priors, prior predictive checks and models with non-informative priors were calculated. For the computation, we used four chains, a thinning rate of two, 5,000 iterations for warm-up, and 10,000 in total. For evaluating the computation, MCMC chain convergence was checked with trace plots and by the R-hat values, and MCMC chain resolution by the effective sampling size (ESS). Posterior predictive checks were done with Kernel density estimates histograms of statistics skew and leave-one-out (LOO) cross-validation [see @gabry2019]. Finally, models were compared with the Bayes factor (BF) and Bayesian *R*² values [@gelman2019]. A non-metric multidimensional scaling ordination (NMDS) with Sørensen dissimilarity was used to visualize variation in species composition in space and time (2D-stress 0.21). Seven species were excluded because they had an accumulated cover over all plots of less than 0.5%. Finally, 343 species were included in the ordination.

@@

Restoration on agricultural soils can have limited success due to high nutrient loads [@walker2004] but mixing with a mineral compnent need not necessarily improve the outcome [@chenot-lescure2022]. Similarly to that study in France, sand admixture reduced nutrient loads and led to higher persistence of sown species while 50% admixture did not further increase the effect. In addition, the effect only appeared in north exposition and the effect size of about 6% in the fourth year after restoration was rather small. The Favorable Conservation Status (FCS) was hardly affected by sand admixture which corresponds to the experiment in a quarry [@chenot-lescure2022]. Substrate depth did not significantly affect persistence or FCS similarly to earlier studies [@baer2004]. Larger differences in soil depths might be necessary to observe negative effects by thicker substrate layers as was shown for prairies [@dornbush2010]. Neither, seed density had an effect on persistence and FCS which fits to results of @Kaulfuss.2022, who found out that a certain amount of seeds are necessary for a successful establishment of target species but higher densities do not further improve the result and have rather a slight negative effect.

The vegetation in south exposition had a different species composition which confirms the findings of @Batori.2016 in Hungary. Here, the differences might have some unwanted experimental reasons: the geotextile, which had been implemented on the southern slope, was removed after two weeks. This led to the drying of many seedlings which was amplified by the intense drought of 2018 and 2019 [@hari2020]. This drought could have decreased germination and establishment of sown species also in general because a moist spring with a mesic temperature improves establishment [@Larson.2021; @orrock2023]. The interesting point here is the lasting negative effect on persistence and FCS which suggests a legacy effect by adverse weather conditions after sowing as it was observed by other studies [@Stuble.2017; @Groves.2020]. The problems during the establishment phase might have led to an alternative trajectory [@suding2004] and could probably only be met by re-sowing. The drying of seedlings might also have leveled the distinction of the seed mixture types on the south exposition.

@@

Our aim was to identify the best seed--substrate combination in respect of restoration effectiveness and biodiversity. For evaluating effectiveness, we measured the persistence of the sown species and for fostering plant biodiversity in general, we measured the FCS. Though, we could not identify an interaction effect for neither of the indices. We would have expected a better performance of hay-meadow seed mixtures with lower sand admixture and for dry grasslands with higher sand admixture. Our results suggest that, at least after four years, the substrate conditions are within the range of both seed mixture types (hay meadows, dry grasslands). Although, both types are clearly plant-sociologically and functionally distinct, they are still relative closely related why they contain shared species and could develop on similar site conditions with modified sub-associations[Appendix A3, @Oberdorfer.1993; @Oberdorfer.1993b; @husicka2003]. Other grassland studies could identify more or less clear interactions of opposing habitat preferences or functional traits along the gradients of productivity, moisture, and nutrients [@Zirbel.2020; @Freitag.2021; @Kaulfuss.2022]. However, these studies did not work with an experimental set up of different seed--substrate combinations, but analyzed the result of habitat and biotic filtering after one, five and about 15 years, respectively. Furthermore, the non-existence of an ideal combination could be explained by priority effects which means that the species of the non-perfect seed mixture could establish earlier and preempted the available niches for the species of related habitat types [@Fukami.2015].

@@

Local setting and design of the multifactorial experiment on grassland sowing at dikes. The experiment was located on a dike at river Danube in SE Germany. The 288 plots were allocated in six blocks (white squares on upper photo) and on the north and south slope (central photo) [Both aerial photos: @bayerischevermessungsverwaltung2023]. Four treatments were conducted: sand admixture, substrate depth, seed density, and seed mixture types H and D (hay meadows, dry grasslands). The western half of a block was with a shallow substrate depth and within this half the substrates had different sand admixtures. The photo on the bottom shows the northern slope of one block in 2021, four years after sowing (Photo: Markus Bauer).

@@

```{r fig.id = "persistence", fig.asp = 0.4848, fig.width = 6.49, fig.cap = "Persistence of sown species of sown grassland communities at a river dike. The plots had substrates with different sand admixtures and were sown with two different seed mixture types. Per plot 20 species were sown. Shown are the medians and 95% credible intervals of the posterior distributions which were derived from a Bayesian linear mixed-effects model (BLMM)."}
knitr::include_graphics(
  here::here("outputs", "figures", "figure_3_persistence_epred_800dpi_16.5x8cm.tiff")
  )
```

@@

```{r fig.id = "fcs", fig.asp = 0.4848, fig.width = 6.49, fig.cap = "The Favorable Conservation Status (FCS) of sown grassland communities at a river dike. The plots had substrates with different sand admixtures and were sown with two different seed mixtures. The zero line indicate that target and non-target species are balanced. Positive values indicate that there are more target species. Shown are the medians and 95% credible intervals of the posterior distributions which were derived from a Bayesian linear mixed-effects model (BLMM)."}
knitr::include_graphics(
  here::here("outputs", "figures", "figure_4_fcs_epred_800dpi_16.5x8cm.tiff")
  )
```

@@

```{r fig.id = "recovery", fig.asp = 0.4848, fig.width = 6.49, fig.cap = "The recovery or successional distance (*d~jt,0~*) of sown grassland communities at a river dike in comparison to the respective reference sites (hay meadows or dry calcareous grasslands). The plots had substrates with different sand admixtures and were sown with two different seed mixtures. The zero lines indicate the mean position of the reference sites for each habitat type on the NMDS axis 1 (Figure 7). The grey area marks the standard deviation of the position of the reference sites. Shown are the medians and 95% credible intervals of the posterior distributions which were derived from a Bayesian linear mixed-effects model (BLMM)."}
knitr::include_graphics(
  here::here("outputs", "figures", "figure_5_recovery_epred_800dpi_16.5x8cm.tiff")
  )
```

@@

The species composition of sown experimental plots on a river dike over time and in comparison with reference sites and the seed mixtures. Both expositions and both seed mixture types are shown in separated panels. The NMDS was based on Sørensen dissimilarity and data of 288 plots which were observed over four years after sowing in 2018 (circles). These experimental plots were compared with the seed mixtures (black squares) and 98 positive and negative reference plots (filled symbols) from older dike grasslands in the surroundings [@bauer2023] and six plots from sPlotOpen [@Sabatini.2021]. The circles show the standard error of the groups. 2D-stress: 0.21.
